---
title: "Assignment 2 Task 1"
author: "Justine Lang"
date: "2023-02-21"
output: 
  html_document: 
    code_folding: hide
---

```{r setup, include=TRUE, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(GGally)
library(broom)
library(AICcmodavg)
library(kableExtra)
library(tidymodels)
```

## Data Overview

Text here


## Data Visualization

Visualizations of differences in height, canopy length, canopy width, and green leaves for *Serenoa repens* and *Sabal etonia*. 

```{r}
#Read in the data

palmetto <- read_csv(here("data", "palmetto.csv"))
```

### Data Wrangling

```{r}
#Rename species codes to species names

palmetto_names <- palmetto %>% 
   mutate(species = case_when(species == "1" ~ "Serenoa repens",
    species == "2" ~ "Sabal etonia"))

#Subset the data

palmetto_subset <- palmetto_names %>% 
  select(species, height:green_lvs) %>% 
  drop_na()

#Convert species to a factor
palmetto_subset$species <- as.factor(palmetto_subset$species)

#class(palmetto_subset$species)
#levels(palmetto_subset$species)
```

### Create Graphs

```{r}
#Explore potential trends

#ggpairs(palmetto_subset, aes(color = species)) 
### Height & length, height & width, green leaves & species look like good relationships to visualize
```

```{r figure 1, fig.align='center', fig.cap="Figure 1: The relationship between plant height (cm) and canopy length (cm) for two palmetto species, observed from 1981 to 2017."}

#Create the first figure

height_length_gg <- ggplot(data = palmetto_subset, aes(x = height, y = length)) +
  geom_point(aes(color = species), show.legend = FALSE) +
  facet_wrap(~species) +
  theme_minimal() +
  labs( x = "Plant Height (cm)",
        y = "Canopy Length (cm)",
        title = "Palmetto Height and Canopy Length") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        axis.title.x = element_text(vjust = 0.2, size = 10),
        axis.title.y = element_text(size = 10),
        axis.text = element_text(size = 8),
        plot.background = element_rect(fill = "white"))+
  scale_color_manual(values = c("green4", "dodgerblue2"))
  
height_length_gg
```

```{r figure 2, fig.align='center', fig.cap="Figure 2: The relationship between plant height (cm) and canopy width (cm) for two palmetto species, observed from 1981 to 2017."}

#Create the second figure

height_width_gg <- ggplot(data = palmetto_subset, aes(x = height, y = width)) +
  geom_point(aes(color = species), show.legend = FALSE) +
  facet_wrap(~species) +
  theme_minimal() +
  labs( x = "Plant Height (cm)",
        y = "Canopy Width (cm)",
        title = "Palmetto Height and Canopy Width") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        axis.title.x = element_text(vjust = 0.2, size = 10),
        axis.title.y = element_text(size = 10),
        axis.text = element_text(size = 8),
        plot.background = element_rect(fill = "white")) +
  scale_color_manual(values = c("green4", "dodgerblue2"))

height_width_gg
```

```{r figure 3, fig.align='center', fig.cap="Figure 3: A histogram comparing the number of green leaves between two palmetto species, observed from 1981 to 2017."}

#Create the third figure

leaves_gg <- ggplot(data = palmetto_subset, aes(x = green_lvs, fill = species)) +
  geom_histogram() +
  facet_wrap(~species) +
  theme_minimal() +
  labs( x = "Number of Green Leaves",
        y = "Count per species",
        title = "Count of Green Leaves Across Palmetto Species") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        axis.title.x = element_text(vjust = 0.2, size = 10),
        axis.title.y = element_text(size = 10),
        axis.text = element_text(size = 8),
        plot.background = element_rect(fill = "white"),
        legend.position = "none")+
  scale_fill_manual(values = c("green4", "dodgerblue2"))

leaves_gg
```


Based on these plots, the easiest way to classify each species would be to count the number of green leaves. There is more overlap between their height, canopy length, and canopy width, though looking at a combination of the variables could further help narrow down classification. 


















