---
title: "Assignment 2 Task 2"
author: "Justine Lang"
date: "2023-02-21"
output: 
  html_document: 
    code_folding: hide
---

```{r setup, include=TRUE, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(kableExtra)
library(Metrics)
```

## Data Overview

This analysis used non-linear least squares (NLS) to estimate parameters of a length to weight
model for lizard populations in New Mexico. The lizard data was collected from pitfall traps located at 11 consumer plots at Jornada Basin LTER site from 1989 to 2006, and includes the variables: species, sex, recapture status, snout-vent length, total length, weight, and whether tail is broken or whole (Lightfoot et al., 2020). This analysis will compare two length to weight NLS models: one for all observed lizard species and one for a subset of the male Western Whiptail lizard (*Cnemidophorus tigrisatus*). 

```{r}
#Read in the data

lizards <- read_csv(here("data", "lizards.csv"))
```


## All Species NLS

\begin{equation}
\hat{W}=a(SVL)^b
\end{equation}


### Select Model

```{r}
#Create a function using given model

func1 <- function(a, b, length){
 out = a *(length^b)
return(out)
}
```


### Initial Guess

```{r}
#Run OLS regression on log transformed data

my_guess_model <- lm(log(lizards$weight) ~ log(lizards$SV_length), data = lizards)

#Get coefficients

#coefficients(my_guess_model)

# (Intercept) log(lizards$SV_length) 
#  -8.256862               2.479417

#Mathematically transform the intercept coefficient to get the guess for parameter a

b <- my_guess_model$coefficients[2]
a <- exp((my_guess_model$coefficients[1]))
```


### Run NLS

```{r}
lizard_nls <- nls(weight ~ func1(a, b, SV_length),
                  data=lizards,
                  start=list(a = a, b = b),
                  trace=FALSE)

#summary(lizard_nls)

broom::tidy(lizard_nls) %>% 
  kable(caption = "Table 1: All Species NLS Results") %>% 
  kable_classic()
```


## Make Prediction

```{r figure 1, fig.align='center', fig.cap="Figure 1: Snout-vent length (mm) and body weight (g) for lizards sampled from Jornada Basin LTER site. Purple points are females and orange are males. The black line shows the predicted data using NLS to estimate parameters."}

#Make a prediction based on the model

lizard_predict <- lizards %>% 
  mutate(predict = predict(lizard_nls,newdata =.))

ggplot(data = lizard_predict) +
  geom_point(aes(x = SV_length,y = weight, color = sex)) +
  labs(x = "Snout-Vent Length (mm)",
      y = "Body Weight (g)",
      title = "NLS Predictions - All Lizard Species")+
  geom_line(aes(x = SV_length,y=predict), color='black') +
  scale_color_manual(values= c("darkmagenta", "orange")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```



## Male Western Whiptail NLS

```{r}
whiptail <- lizards %>% 
  filter(spp == "CNTI", sex == "M")
```


### Select Model

```{r}
#Create a function using given model (already completed)

func1 <- function(a, b, length){
 out= a*(length^b)
return(out)
}
```


### Initial Guess

```{r}
#Run OLS regression on log transformed data

my_guess_model_whip <- lm(log(whiptail$weight) ~ log(whiptail$SV_length), data = whiptail)

#Get coefficients

#coefficients(my_guess_model_whip)

# (Intercept) log(lizards$SV_length) 
#  -9.023963             2.698108

#Mathematically transform the intercept coefficient to get the guess for parameter a

b_whip <- my_guess_model_whip$coefficients[2]
a_whip <- exp((my_guess_model_whip$coefficients[1]))
```


### Run NLS

```{r}
whiptail_nls <- nls(weight ~ func1(a, b, SV_length),
                  data = whiptail,
                  start = list(a = a_whip, b = b_whip),
                  trace = FALSE)

#summary(whiptail_nls)

broom::tidy(whiptail_nls) %>% 
  kable(caption = "Table 2: Male Western Whiptail NLS Results") %>% 
  kable_classic()
```


## Make Prediction

```{r figure 2, fig.align='center', fig.cap="Figure 2: Snout-vent length (mm) and body weight (g) for male Western Whiptail lizards. The orange line represents the NLS predictions for the male Western Whiptail subset, and the black line represents the NLS predictions for all species and sexes. The root-mean-square error (RMSE) for the subset model is 3.35, and the RMSE for the all species model is 3.46. The lower RMSE indicates the male Western Whiptail model fits the subset data better than the all species model."}

#Prediction for whiptail model

whiptail_predict <- whiptail %>% 
  mutate(predict = predict(whiptail_nls, newdata=.)) %>% 
  
#Prediction for lizard model on whiptail dataset
  
  mutate(predict_all = predict(lizard_nls, newdata=.))

#Plot both predictions on graph together

ggplot(data = whiptail_predict) +
  geom_point(aes(x = SV_length, y = weight)) +
  labs(x = "Snout-Vent Length (mm)",
      y = "Body Weight (g)",
      title = "NLS Predictions - Western Whiptail Comparison") +
    geom_line(aes(x = SV_length,y = predict, color = "male")) +
  geom_line(aes(x = SV_length, y = predict_all, color = "all")) +
  scale_colour_manual(name= "legend", values=c("male"="orange","all"= "black")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

#Calculate RMSE for both models

rmsewhiptail <- modelr::rmse(whiptail_nls, whiptail)
rmselizard <- modelr::rmse(lizard_nls, whiptail)
```


## Citation

Lightfoot, D. and W.G. Whitford. 2020. Lizard pitfall trap data from 11 NPP study locations at the Jornada Basin LTER site, 1989-2006 ver 37. Environmental Data Initiative. https://doi.org/10.6073/pasta/4a6e258fb49c31e222ecbbcfd128967f



